---
title: "Filtering: Data-Driven"
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{filtering_data_driven}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk: 
    collapse: true
    comment: '#>'
    fig.align: "center"
---


## Preliminary


```{r warning=FALSE, message=FALSE, results='hide', fig.show='hide'}
## load R package
library(msDiaLogue)
## preprocessing
fileName <- "../inst/extdata/Toy_Spectronaut_Data.csv"
dataSet <- preprocessing(fileName,
                         filterNaN = TRUE, filterUnique = 2,
                         replaceBlank = TRUE, saveRm = TRUE)
## transformation
dataTran <- transform(dataSet, logFold = 2)
## annotation-based filtering
dataFiltAnno <- filterOutIn(dataTran, listName = "ALBU_BOVIN",
                            removeList = TRUE, saveRm = TRUE)
## normalization
dataNorm <- normalize(dataFiltAnno, normalizeType = "median")
```


## Examples


Before imputing missing (`NA`) values, proteins can be filtered based on
the proportion of non-missing observations using `filterNA()`.


The `minProp` argument sets the minimum required non-missing proportion for
a protein to be retained, and the `by` argument determines whether this
proportion is evaluated within each condition (`by = "cond"`) or across all
samples (`by = "all"`).


For example, setting `minProp = 0.51` and `by = "cond"` requires that a protein
must be observed in at least 51\% of replicates within each condition.
If a protein fails to meet the threshold in any condition, it will be removed.
And `saveRm = TRUE` indicates that the filtered data will be saved as a `.csv`
file named *filtered_NA_data.csv* in the current working directory.


```{r}
dataFilt <- filterNA(dataNorm, minProp = 0.51, by = "cond", saveRm = TRUE)
```


<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataFilt)
```
</div>


<div class="note">
**Note:** There is no rule in the field of proteomics for filtering based on
percentage of missingness, similar to there being no rule for the number of
replicates required to draw a conclusion. However, reproducible observations
make conclusions more credible. Setting the `minProp` to 0.51 requires that
any protein be observed in a majority of the replicates by condition in order to
be considered. For 3 replicates, this would require 2 measurements to allow
imputation of the 3rd value. If only 1 measurement is seen, the other values
will remain NA, and will be filtered out in a subsequent step.  </div>


<div class="nav-button left-button">
<a href="normalization.html">← Previous</a>
</div>
<div class="nav-button right-button">
<a href="imputation.html">Next →</a>
</div>

