---
title: "Filtering: Annotation-Based"
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{filtering_annotation_based}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk: 
    collapse: true
    comment: '#>'
    fig.align: "center"
---


## Preliminary


```{r warning=FALSE, message=FALSE, results='hide', fig.show='hide'}
## load R package
library(msDiaLogue)
## preprocessing
fileName <- "../inst/extdata/Toy_Spectronaut_Data.csv"
dataSet <- preprocessing(fileName,
                         filterNaN = TRUE, filterUnique = 2,
                         replaceBlank = TRUE, saveRm = TRUE)
## transformation
dataTran <- transform(dataSet, logFold = 2)
```


## Examples


### Case 1. Remove proteins specified by the user in this step and keep everything else.


In the example below, the specific protein with the identifier `"ALBU_BOVIN"`
will be removed. If `removeList = TRUE`, this function will remove what you've
specified and keep the rest.


```{r}
dataFiltAnno <- filterOutIn(dataTran, listName = "ALBU_BOVIN",
                            removeList = TRUE, saveRm = TRUE)
```


<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataFiltAnno)
```
</div>


If `saveRm = TRUE`, the filtered-out data (`"ALBU_BOVIN"`) will be saved as
an `.xlsx` file named `filterOutIn.xlsx` in the current working directory, and
you can inspect this list to see what was removed.


To remove multiple entries, users may combine exact matching `listName` with
pattern matching `regexName`. For example, in addition to removing `"ALBU_BOVIN"`,
the following code also removes any protein whose identifier contains the string
`"HUMAN"`.


```{r}
dataFiltAnno2 <- filterOutIn(dataTran,
                             listName = "ALBU_BOVIN", regexName = "HUMAN",
                             removeList = TRUE, saveRm = TRUE)
```


<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataFiltAnno2)
```
</div>


### Case 2. Keep the proteins specified by the user in this step and remove everything else.


If we set `removeList` to FALSE, running this code will remove everything you
DIDN'T specify and keep only things that matched your search terms.


```{r}
dataFiltAnno3 <- filterOutIn(dataTran,
                             listName = "ALBU_BOVIN", regexName = "HUMAN",
                             removeList = FALSE)
```


<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataFiltAnno3)
```
</div>


### Extension


Besides protein names, the function `filterOutIn()` can filter proteins using
additional protein information.


+ For Spectronaut: `"PG.Genes"`, `"PG.ProteinAccession"`,
  `"PG.ProteinDescriptions"`, and `"PG.ProteinName"`.


+ For Scaffold: `"ProteinDescriptions"`, `"AccessionNumber"`, and `"AlternateID"`.


```{r}
dataFiltAnno4 <- filterOutIn(dataTran,
                             listName = c("Putative zinc finger protein 840",
                                          "Bovine serum albumin"),
                             by = "PG.ProteinDescriptions",
                             removeList = FALSE)
```


In this case, the proteins whose `"PG.ProteinDescriptions"` match with
`"Putative zinc finger protein 840"` or `"Bovine serum albumin"` will be kept.


<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataFiltAnno4)
```
</div>


## Details


In some cases, a researcher may wish to filter out a specific protein or
proteins from the dataset. The most common instance of this would be proteins
identified from the common contaminants database, where we don't want something
like BSA to be matched to a human protein because the search algorithm didn't
have the correct option available, but we don't actually care about BSA itself
and want to leave it out of our visualization.  Other examples may be filtering
out entries from the decoy database (specific to a Scaffold file only, will
not be present in a Spectronaut file), or a mixed-species experiment where the
researcher wants to evaluate data from only one species at a time. This step
allows you to set aside specific proteins from downstream analysis, using either
an exact match identifier (the `listName` argument), or text-containing
identifiers (the `regexName` argument).


<div class="note">
**Note:** If both `listName` and `regexName` are defined, the proteins to be
selected or removed is the union of the two terms. \
**Keep in mind:** Removal of any proteins, including common contaminants, will
affect any global calculations performed after this step (such as normalization).
This should not be done without a clear understanding of how this will affect
your results. </div>


<div class="nav-button left-button">
<a href="transformation.html">← Previous</a>
</div>
<div class="nav-button right-button">
<a href="normalization.html">Next →</a>
</div>

