---
title: "Imputation"
output: rmarkdown::html_vignette
author: 
  - |
    Shiying Xiao$^1$, Charles Watt$^1$, Jennifer C. Liddle$^2$, Jeremy L. Balsbaugh$^2$, Timothy E. Moore$^3$
  - |
    $^1$Department of Statistics, UConn \
    $^2$Proteomics and Metabolomics Facility, UConn \
    $^3$Statistical Consulting Services, UConn
date: "`r format(file.info('imputation.Rmd')$mtime, '%Y-%m-%d')`"
bibliography:
  - ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{imputation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
```


## Preliminary


```{r warning=FALSE, message=FALSE, results='hide', fig.show='hide'}
## load R package
library(msDiaLogue)
## preprocessing
fileName <- "../inst/extdata/Toy_Spectronaut_Data.csv"
dataSet <- preprocessing(fileName,
                         filterNaN = TRUE, filterUnique = 2,
                         replaceBlank = TRUE, saveRm = TRUE)
## transformation
dataTran <- transform(dataSet, logFold = 2)
## normalization
dataNorm <- normalize(dataTran, normalizeType = "quant")
```


## Examples


For example, to impute the NA value of `dataNorm` using `impute.min_local()`,
set the required percentage of values that must be present in a given protein by
condition combination for values to be imputed to 51\%.


<div class="note">
**Note:** There is no rule in the field of proteomics for filtering based on
percentage of missingness, similar to there being no rule for the number of
replicates required to draw a conclusion. However, reproducible observations
make conclusions more credible. Setting the `reqPercentPresent` to 0.51 requires
that any protein be observed in a majority of the replicates by condition in
order to be considered. For 3 replicates, this would require 2 measurements to
allow imputation of the 3rd value. If only 1 measurement is seen, the other
values will remain NA, and will be filtered out in a subsequent step.  </div>


```{r}
dataImput <- impute.min_local(dataNorm, reportImputing = FALSE,
                              reqPercentPresent = 0.51)
```


<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataImput)
```
</div>


If `reportImputing = TRUE`, the returned result structure will be altered to a
list, adding a shadow data frame with imputed data labels, where 1 indicates the
corresponding entries have been imputed, and 0 indicates otherwise.


After the above imputation, any entries that did not pass the percent present
threshold will still have NA values and will need to be filtered out.


```{r}
dataImput <- filterNA(dataImput, saveRm = TRUE)
```


where `saveRm = TRUE` indicates that the filtered data will be saved as a .csv
file named *filtered_NA_data.csv* in the current working directory.


The `dataImput` is as follows:


<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataImput)
```
</div>


## Details


The two primary MS/MS acquisition types implemented in large scale MS-based
proteomics have unique advantages and disadvantages. Traditional Data-Dependent
Acquisition (DDA) methods favor specificity in MS/MS sampling over comprehensive
proteome coverage. Small peptide isolation windows (<3 m/z) result in MS/MS
spectra that contain fragmentation data from ideally only one peptide. This
specificity promotes clear peptide identifications but comes at the expense of
added scan time. In DDA experiments, the number of peptides that can be selected
for MS/MS is limited by instrument scan speeds and is therefore prioritized by
highest peptide abundance.  Low abundance peptides are sampled less frequently
for MS/MS and this can result in variable peptide coverage and many missing
protein data across large sample datasets.


Data-Independent Acquisition (DIA) methods promote comprehensive peptide
coverage over specificity by sampling many peptides for MS/MS simultaneously.
Sequential and large mass isolation windows (4-50 m/z) are used to isolate large
numbers of peptides at once for concurrent MS/MS. This produces complicated
fragmentation spectra, but these spectra contain data on every observable
peptide.  A major disadvantage with this type of acquisition is that DIA MS/MS
spectra are incredibly complex and difficult to deconvolve.  Powerful and
relatively new software programs like Spectronaut are capable of successfully
parsing out which fragment ions came from each co-fragmented peptide using
custom libraries, machine learning algorithms, and precisely determined
retention times or measured ion mobility data. Because all observable ions are
sampled for MS/MS, DIA reduces missingness substantially compared to DDA, though
not entirely.


Various imputation methods have been developed to address the missing-value
issue and assign a reasonable guess of quantitative value to proteins with
missing values. So far, this package provides 10 imputation methods for use:


1. `impute.min_local()`: Replaces missing values with the lowest measured value
for that protein in that condition.

2. `impute.min_global()`: Replaces missing values with the lowest measured value
from any protein found within the entire dataset.

3. `impute.knn()`: Replaces missing values using the k-nearest neighbors
algorithm [@troyanskaya2001missing].

4. `impute.knn_seq()`: Replaces missing values using the sequential k-nearest
neighbors algorithm [@kim2004reuse].

5. `impute.knn_trunc()`: Replaces missing values using the truncated k-nearest
neighbors algorithm [@shah2017distribution].

6. `impute.nuc_norm()`: Replaces missing values using the nuclear-norm
regularization [@hastie2015matrix].

7. `impute.mice_cart()`: Replaces missing values using the classification and
regression trees [@breiman1984classification; @doove2014recursive; @van2018flexible].

8. `impute.mice_norm()`: Replaces missing values using the Bayesian linear
regression [@rubin1987multiple; @schafer1997analysis; @van2011mice].

9. `impute.pca_bayes()`: Replaces missing values using the Bayesian principal
components analysis [@oba2003bayesian].

10. `impute.pca_prob()`: Replaces missing values using the probabilistic
principal components analysis [@stacklies2007pcamethods].

Additional methods will be added later.


## Reference {-}


<div class="nav-button left-button">
<a href="normalization.html">← Previous</a>
</div>
<div class="nav-button right-button">
<a href="summarization.html">Next →</a>
</div>

