---
title: "Normalization"
output: rmarkdown::html_vignette
author: 
  - |
    Shiying Xiao$^1$, Charles Watt$^1$, Jennifer C. Liddle$^2$, Jeremy L. Balsbaugh$^2$, Timothy E. Moore$^3$
  - |
    $^1$Department of Statistics, UConn \
    $^2$Proteomics and Metabolomics Facility, UConn \
    $^3$Statistical Consulting Services, UConn
date: "`r format(file.info('normalization.Rmd')$mtime, '%Y-%m-%d')`"
bibliography:
  - ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{normalization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
```


## Preliminary


```{r warning=FALSE, message=FALSE, results='hide', fig.show='hide'}
## load R package
library(msDiaLogue)
## preprocessing
fileName <- "../tests/testData/Toy_Spectronaut_Data.csv"
dataSet <- preprocessing(fileName,
                         filterNaN = TRUE, filterUnique = 2,
                         replaceBlank = TRUE, saveRm = TRUE)
## transformation
dataTran <- transform(dataSet, logFold = 2)
```


## Example


```{r}
dataNorm <- normalize(dataTran, normalizeType = "quant")
```


The message "Warning: Removed 55 rows containing non-finite values" indicates
the presence of 55 NA (Not Available) values in the data. These NA values arise
when a protein was not identified in a particular sample or condition and are
automatically excluded when generating the boxplot but retained in the actual
dataset.


<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataNorm)
```
</div>


## Details


Normalization is designed to address systematic biases in the data. Biases can
arise from inadvertent sample grouping during generation or preparation, from
variations in instrument performance during acquisition, analysis of different
peptide amounts across experiments, or other reasons. These factors can
artificially mask or enhance actual biological changes. 


Many normalization methods have been developed for large datasets, each with its
own strengths and weaknesses. The following factors should be considered when
choosing a normalization method:


a) Experiment-Specific Normalization: \
Most experiments run with [UConn PMF](https://proteomics.uconn.edu) are
normalized by injection amount at the time of analysis to facilitate comparison.
"Amount" is measured by UV absorbance at 280 nm, a standard method for generic
protein quantification.


b) Assumption of Non-Changing Species: \
Most biological experiments implicitly assume that the majority of measured
species in an experiment will not change across conditions. This assumption is
more robust the more measurements your experiment has (e.g. several thousand
proteins). It may not be true at all for small datasets (tens of proteins).


If you are analyzing a batch of samples with very different complexities (e.g.
a set of IPs where the control samples have tens of proteins and the
experimental samples have hundreds of proteins), you should not normalize all of
these together, but break them up into subsets of similar complexity.


By default, normalization is performed across samples, adjusting protein
expression levels within each sample relative to the other samples.
So far, this package provides eight normalization methods for use:


1. "auto": Auto scaling (mean centering and then dividing by the standard
deviation of each variable) [@jackson1991user].

2. "level": Level scaling (mean centering and then dividing by the mean of each variable).

3. "mean": Mean centering.

4. "median": Median centering.

5. "pareto": Pareto scaling (mean centering and then dividing by the square root
of the standard deviation of each variable).

6. "quant": Quantile normalization [@bolstad2003comparison].

7. "range": Range scaling (mean centering and then dividing by the range of each
variable).

8. "vast": Variable stability (VAST) scaling [@keun2003improved].


## Reference {-}


<div class="nav-button left-button">
<a href="filtering.html">← Previous</a>
</div>
<div class="nav-button right-button">
<a href="imputation.html">Next →</a>
</div>

