---
title: "Preprocessing"
output: rmarkdown::html_vignette
author: 
  - |
    Shiying Xiao$^1$, Charles Watt$^1$, Jennifer C. Liddle$^2$, Jeremy L. Balsbaugh$^2$, Timothy E. Moore$^3$
  - |
    $^1$Department of Statistics, UConn \
    $^2$Proteomics and Metabolomics Facility, UConn \
    $^3$Statistical Consulting Services, UConn
date: "`r format(file.info('preprocessing.Rmd')$mtime, '%Y-%m-%d')`"
bibliography:
  - ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{preprocessing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
```


## Preliminary


```{r warning=FALSE, message=FALSE}
## load R package
library(msDiaLogue)
```


## Example


```{r eval=FALSE}
## if the raw data is in a .csv file
fileName <- "../inst/extdata/Toy_Spectronaut_Data.csv"
dataSet <- preprocessing(fileName,
                         filterNaN = TRUE, filterUnique = 2,
                         replaceBlank = TRUE, saveRm = TRUE)
```


<div class="note">
**Note:** `preprocessing()` does not perform a transformation on your data. You
still need to use the function `transform()`. </div>


```{r}
## if the raw data is in an .Rdata file
load("../inst/extdata/Toy_Spectronaut_Data.RData")
dataSet <- preprocessing(dataSet = Toy_Spectronaut_Data,
                         filterNaN = TRUE, filterUnique = 2,
                         replaceBlank = TRUE, saveRm = TRUE)
```


<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataSet, align = "l")
```
</div>


## Details


The function `preprocessing()` takes a `.csv` file of summarized protein
abundances, exported from **Spectronaut**. The most important columns that need
to be included in this file are: `R.Condition`, `R.Replicate`,
`PG.ProteinAccessions`, `PG.ProteinNames`, `PG.NrOfStrippedSequencesIdentified`,
and `PG.Quantity`. This function will reformat the data and provide
functionality for some initial filtering (based on the number of unique peptides).
The steps below describe the functions that happen in the Preprocessing code.


**1\.** Loads the raw data


+ If the raw data is in a .csv file
[Toy_Spectronaut_Data.csv](https://github.com/uconn-scs/msDiaLogue/blob/main/inst/extdata/Toy_Spectronaut_Data.csv),
specify the `fileName` to read the raw data file into **R**.


+ If the raw data is stored as an .RData file
[Toy_Spectronaut_Data.RData](https://github.com/uconn-scs/msDiaLogue/blob/main/inst/extdata/Toy_Spectronaut_Data.RData),
first load the data file directly, then specify the `dataSet` in the function.


**2\.** Filters out identified proteins that exhibit "NaN" quantitative values


NaN, which stands for 'Not a Number,' can be found in the PG.Quantity column for
proteins that were identified by MS and MS/MS evidence in the raw data, but all
peptides from that protein lack an associated integrated peak area or intensity.
This usually occurs in low abundance peptides that exhibit intensities close to
the limit of detection resulting in poor signal-to-noise (S/N) and/or when there
is interference from other co-eluting peptide ions with very similar or
identical m/z values that lead to difficulty in parsing out individual intensity
profiles.


**3\.** Applies a unique peptides per protein filter


General practice in the proteomics field is to filter out proteins which were
identified on the basis of a single peptide. Because approximately 1\% of all
identified peptides are false positive matches, it's more likely that 1 peptide
was incorrectly identified and that protein ID is incorrect than that, for
example, 5 peptides from the same protein were all incorrectly identified and
that protein ID is incorrect. We recommend focusing on proteins with 2 or more
peptide identifications, as these will be higher confidence. If you have a
protein of interest with only 1 peptide identified, contact [PMF faculty](https://proteomics.uconn.edu/about-us/)
and we can help you evaluate the evidence from the raw data to determine
believability.


**4\.** Adds accession numbers to identified proteins without informative names


**Spectronaut** reports contain 4 different columns of identifying information:


+ `PG.Genes`, which is the gene name (e.g. CDK1).
+ `PG.ProteinAccessions`, which is the UniProt identifier number for a unique
entry in the online database (e.g. P06493).
+ `PG.ProteinDescriptions`, which is the protein name as provided on UniProt
(e.g. cyclin-dependent kinase 1).
+ `PG.ProteinNames`, which is a concatenation of an identifier and the species
(e.g. CDK1_HUMAN).


Every entry in UniProt will have an accession number, but may not have all of
the other identifiers, due to incomplete annotation.  Because Uniprot includes
entries for fragments of proteins and some proteins entries are redundant, a
peptide can match to multiple entries for the same protein, which generates
multiple possible identifiers in **Spectronaut**. Further, the `ProteinNames`
entry in **Spectronaut** can switch formats: the preference is accession number
and species, but can also be gene name and species instead.


This option tells **msDiaLogue** to substitute the accession number for an
identifier if it tries to pull an identifier from a column with no information.


<div class="note">
**Note:** Not all proteins can be identified unambiguously. In many cases, the
identified peptides can be found in multiple protein sequences, which yields a
protein group or protein cluster rather than a single protein identification.
When this happens, the accession numbers for all potential matches are
concatenated into one string, separated by periods. When you see long strings of
multiple identifiers later in your data processing, this is why. **Spectronaut**
sorts these alphanumerically, so you should not assume that the first protein in
the list is most likely to be correct (other search algorithms such as MaxQuant,
which is used in PMF for most Scaffold-based results, do rank protein
cluster IDs by likelihood of correctness). </div>


**5\.** Saves a document to your working directory with all filtered out data,
if desired


If `saveRm = TRUE`, the data removed in step 2 (*preprocess_filterNaN.csv*) and
step 3 (*preprocess_filterUnique.csv*) will be saved in the current working
directory.


As part of the `preprocessing()`, a histogram of $log_2$-transformed protein
abundances is provided. This is a helpful way to confirm that the data have been
read in correctly, and there are no issues with the numerical values of the
protein abundances. Ideally, this histogram will appear fairly symmetrical
(bell-shaped) without too much skew towards smaller or larger values.


<div class="nav-button right-button">
<a href="transformation.html">Next →</a>
</div>

