---
title: "Import Scaffold Report for msDiaLogue"
output: rmarkdown::html_vignette
author: 
  - |
    Shiying Xiao$^1$, Charles Watt$^1$, Jennifer C. Liddle$^2$, Jeremy L. Balsbaugh$^2$, Timothy E. Moore$^3$
  - |
    $^1$Department of Statistics, UConn \
    $^2$Proteomics and Metabolomics Facility, UConn \
    $^3$Statistical Consulting Services, UConn
date: "`r Sys.Date()`"
bibliography:
  - ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{A Usage Template}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
```

The default data used in msDiaLogue is from **Spectronaut**; here, we introduce the import
from **Scaffold**.

The function `preprocessing_scaffold()` takes a `.xls` file of preprocessed protein
abundances, exported from **Scaffold**. This function will reformat the data.

+ If the raw data is in a .xls file
[Toy_Scaffold_Data.xls](https://github.com/uconn-scs/msDiaLogue/blob/main/tests/testData/Toy_Scaffold_Data.xls),
specify the `fileName` to read the raw data file into **R**.

+ If the raw data is stored as an .RData file
[Toy_Scaffold_Data.RData](https://github.com/uconn-scs/msDiaLogue/blob/main/tests/testData/Toy_Scaffold_Data.RData),
first load the data file directly, then specify the `dataSet` in the function.

**NOTE:** 2 peptide filter and normalization and abundance definition would have to be set
by user in Scaffold, but decoys and contaminants could be removed on the backend - they'll
be flagged in the protein name.

```{r warning=FALSE, message=FALSE}
library(msDiaLogue)
```

```{r eval=FALSE}
## if the raw data is in a .xls file
dataSet <- preprocessing_scaffold(fileName = "../tests/testData/Toy_Scaffold_Data.xls")
```

```{r}
## if the raw data is in an .Rdata file
load("../tests/testData/Toy_Scaffold_Data.RData")
dataSet <- preprocessing_scaffold(dataSet = Toy_Scaffold_Data)
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataSet, align = "l")
```
</div>