---
title: "Instructions for Exporting Data from Scaffold for Use in msDiaLogue Analysis"
output: rmarkdown::html_vignette
author: 
  - |
    Shiying Xiao$^1$, Charles Watt$^1$, Jennifer C. Liddle$^2$, Jeremy L. Balsbaugh$^2$, Timothy E. Moore$^3$
  - |
    $^1$Department of Statistics, UConn \
    $^2$Proteomics and Metabolomics Facility, UConn \
    $^3$Statistical Consulting Services, UConn
date: "`r format(file.info('scaffold.Rmd')$mtime, '%Y-%m-%d')`"
bibliography:
  - ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{scaffold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
```


This script is to allow PMF users to apply msDiaLogue functions to proteomics
results provided in **Scaffold 5**. msDiaLogue was originally developed to
provide an **R**-based analysis workflow for DIA data searched in **Spectronaut**
(which could not be uploaded into **Scaffold 5** and thus left PMF users with
results in the form of crashingly large Excel files before msDiaLogue became
available). The msDiaLogue initial step, Preprocessing, was built for a
**Spectronaut** report format. This expansion gives msDiaLogue the capability to
preprocess data from a **Scaffold** report format.


Future additions to accommodate report formats from other search algorithms or
softwares are planned, but not available at this time.


To generate an input report for msDiaLogue from **Scaffold**, first
<span style="color:rgb(49,108,244);">
adjust the following 6 parameters in your **Scaffold 5** file</span>.


### 1. Define Quantitative Value by an Intensity Measurement in Scaffold


The quantitative value **must** be set to an intensity measurement and not a
count measurement. Inside Scaffold 5, on the Samples tab, the Display Options
dropdown menu at the top left of the window must be set to "Quantitative Value"
and the quantitative value must be defined (in Experiment --> Quantitative
Analysis --> Other Settings, Quantitative Method dropdown). PMF recommends the
Average Precursor Intensity value.


<div class="note">
**Note:** Any value may be selected here, if you prefer a different
quantification schema, with the exception of Total Spectra or Weighted Spectra.
Do not use these. </div>


### 2. Turn Off Normalization and All Quantitative Comparisons in Scaffold


Normalization **must** be turned off and **no** quantitative analyses should be
applied. In (Experiment --> Quantitative Analysis
--> Other Settings), make sure the "Use Normalization" box is unchecked. You
will have the option to normalize by various methods in msDiaLogue, but you
should not stack normalizations between programs. Also make sure that the
"Statistical Test" pane (upper left) in the "Quantitative Analysis" window is
set to "No Test" and the "Fold Change" pane (lower left) is set to
"No Fold Change".


<div class="note">
**Note:** Non-quantifiable abundance values are reported in **Scaffold** as 0
(not identified or not quantified) or 1 (identified but not quantified), and
msDiaLogue `preprocessing` knows to treat these as NA via `zeroNA = TRUE` and
`oneNA = TRUE`. If normalization is applied in **Scaffold**, non-quantifiable
abundances are transformed into fractional values which will not be converted
properly in `preprocessing` and your data will no longer yield sensible analyses. </div>


### 3. Confirm Experiment has Required Minimum Replicates and Conditions


The experiment **must** contain a minimum of 2 conditions, and each condition
**must** have a minimum of 3 replicates. More conditions are fine, more
replicates are fine, and conditions do not need to have the same number of
replicates. Having fewer than 3 replicates for any condition, or having only 1
condition, will throw an error in msDiaLogue and you will not be able to process
your data.


<div class="note">
**Note:** You can't sensibly estimate sample variance from 2 replicates, so
statistical tests really should not be run on fewer than 3 replicates in general,
whether in msDiaLogue or other programs. </div>


### 4. Confirm or Correct Sample Name Formatting


The samples **must** be named in the following format:
YYYYMMDD_initials_condition_replicate# (e.g. 20240101_JL_ctrl_1). Your files may
already be named this way; please check in **Scaffold** in the "Samples" tab,
for the heading above any quantitative value column. The sample name will appear
vertically. If the name is not formatted as above, you can change it by going
into the "Load Data" tab, selecting the tab of each sample individually,
right-clicking the tab, choosing "Edit BioSample", typing the correct name
format into the "Sample Name" box, and clicking "Apply".


<div class="note">
**Note:** **Scaffold 5** will send you back to the "Samples" tab each time
"Apply" is clicked, so this step is a little tedious to do in **Scaffold**. You
can also fix this in **Excel** after exporting, before you import into **R**,
which will likely be easier. </div>


### 5. Filter Dataset for Proteins Identified by >1 Peptide


We **strongly recommend** that you filter the dataset to hide proteins that only
had 1 peptide identified. In the "Samples" tab, at the top-most menu bar, in the
"Min # Peptides" dropdown, set this to 2.


<div class="note">
**Note:** This is not required, and if you choose to evaluate the dataset with
1-peptide identifications included, msDiaLogue will not stop you. However, the
**Scaffold** report does not provide information about how many peptides/protein
were identified, so unlike with a **Spectronaut**-based report, you cannot
filter based on this information once in msDiaLogue. </div>


### 6. Ensure No Protein Clusters are Expanded in Samples Tab


All protein clusters **must** be collapsed. In the "Samples" tab, in the very
first column (header "#"), right click any of the numbered entries here, select
"Clusters", and select "Collapse All". This converts any gray-shaded entries,
where multiple proteins are inferred from the peptide evidence, from an expanded
view with a line for each possible protein match in that cluster to a collapsed
view with only the first entry appearing in the table. In msDiaLogue, this first
entry and its accession number is the only one that will be represented in your
data from this cluster.


<hr>


You're now ready to export the data. From the "Samples" tab in **Scaffold**,
right-click anywhere in the main data table, choose "Export" (bottom of menu),
and "Export to Excel".


Save with a descriptive filename that will make sense to someone else in the
future and choose the location you'll be using as your working directory in **R**.


The report can be saved as .xls or .csv.


You may now use the `preprocessing_scaffold` script available on this page and
pick up at the `transformation` step in the msDiaLogue script as provided in the
main [Usage Template](https://uconn-scs.github.io/msDiaLogue/articles/usage_template.html) page.

<div class="note">
**Note:** You cannot use the function "preprocessing" in the main template, as that is designed for 
inputs in a Spectronaut report format.  The function for Scaffold data is called "preprocessing_Scaffold",
given in the code box below. </div>


+ If the raw data is in a .xls file
[Toy_Scaffold_Data.xls](https://github.com/uconn-scs/msDiaLogue/blob/main/tests/testData/Toy_Scaffold_Data.xls),
specify the `fileName` to read the raw data file into **R**.

+ If the raw data is stored as an .RData file
[Toy_Scaffold_Data.RData](https://github.com/uconn-scs/msDiaLogue/blob/main/tests/testData/Toy_Scaffold_Data.RData),
first load the data file directly, then specify the `dataSet` in the function.


<div class="note">
**Note:** Decoys and contaminants can be removed either in the **Excel** report
before Preprocessing, or can be removed in msDiaLogue with the filter step, if
the accession number has the "CON__" or "DECOY__" prefix. (This prefix is not
applied with search algorithms like MSFragger, so alternative filters would need
to be developed in that case.) </div>


```{r warning=FALSE, message=FALSE}
library(msDiaLogue)
```


```{r eval=FALSE}
## if the raw data is in a .xls file
dataSet <- preprocessing_scaffold(fileName = "../tests/testData/Toy_Scaffold_Data.xls",
                                  zeroNA = TRUE, oneNA = TRUE)
```


```{r}
## if the raw data is in an .Rdata file
load("../tests/testData/Toy_Scaffold_Data.RData")
dataSet <- preprocessing_scaffold(dataSet = Toy_Scaffold_Data,
                                  zeroNA = TRUE, oneNA = TRUE)
```


<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataSet, align = "l")
```
</div>

