---
title: "Summarization"
output: rmarkdown::html_vignette
author: 
  - |
    Shiying Xiao$^1$, Charles Watt$^1$, Jennifer C. Liddle$^2$, Jeremy L. Balsbaugh$^2$, Timothy E. Moore$^3$
  - |
    $^1$Department of Statistics, UConn \
    $^2$Proteomics and Metabolomics Facility, UConn \
    $^3$Statistical Consulting Services, UConn
date: "`r format(file.info('summarization.Rmd')$mtime, '%Y-%m-%d')`"
bibliography:
  - ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{summarization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
```


## Preliminary


```{r warning=FALSE, message=FALSE, results='hide', fig.show='hide'}
## load R package
library(msDiaLogue)
## preprocessing
fileName <- "../tests/testData/Toy_Spectronaut_Data.csv"
dataSet <- preprocessing(fileName,
                         filterNaN = TRUE, filterUnique = 2,
                         replaceBlank = TRUE, saveRm = TRUE)
## transformation
dataTran <- transform(dataSet, logFold = 2)
## normalization
dataNorm <- normalize(dataTran, normalizeType = "quant")
## imputation
dataImput <- impute.min_local(dataNorm, reportImputing = FALSE,
                              reqPercentPresent = 0.51)
## filtering
dataImput <- filterNA(dataImput, saveRm = TRUE)
```


## Example


```{r}
dataSumm <- summarize(dataImput, saveSumm = TRUE)
```


<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataSumm)
```
</div>


## Details


This summarization provides a table of values for each protein in the final
dataset that include the final processed abundances and fold changes in each
condition, and that protein's statistical relation to the global dataset in
terms of its mean, median, standard deviation, and other parameters.


The column "Stat" in the generated result includes the following statistics:

+ n: Number.
+ mean: Mean.
+ sd: Standard deviation.
+ median: Median.
+ trimmed: Trimmed mean with a trim of 0.1.
+ mad: Median absolute deviation (from the median).
+ min: Minimum.
+ max: Maximum.
+ range: The difference between the maximum and minimum value.
+ skew: Skewness.
+ kurtosis: Kurtosis.
+ se: Standard error.


<div class="nav-button left-button">
<a href="imputation.html">← Previous</a>
</div>
<div class="nav-button right-button">
<a href="analysis.html">Next →</a>
</div>

