---
title: "Summarization"
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{summarization}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk: 
    collapse: true
    comment: '#>'
    fig.align: "center"
---


## Preliminary


```{r warning=FALSE, message=FALSE, results='hide', fig.show='hide'}
## load R package
library(msDiaLogue)
## preprocessing
fileName <- "../inst/extdata/Toy_Spectronaut_Data.csv"
dataSet <- preprocessing(fileName,
                         filterNaN = TRUE, filterUnique = 2,
                         replaceBlank = TRUE, saveRm = TRUE)
## transformation
dataTran <- transform(dataSet, logFold = 2)
## annotation-based filtering
dataFiltAnno <- filterOutIn(dataTran, listName = "ALBU_BOVIN",
                            removeList = TRUE, saveRm = TRUE)
## normalization
dataNorm <- normalize(dataFiltAnno, normalizeType = "median")
## data-driven filtering
dataFilt <- filterNA(dataNorm, minProp = 0.51, by = "cond", saveRm = TRUE)
## imputation
dataImput <- impute.min_local(dataFilt)
```


## Example


```{r}
dataSumm <- summarize(dataImput, saveSumm = TRUE)
```


<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataSumm)
```
</div>


## Details


The `summarization()` provides a table of values for each protein in the final
dataset that include the final processed abundances and fold changes in each
condition, and that protein's statistical relation to the global dataset
in terms of its mean, median, standard deviation, and other parameters.


The column `"Stat"` in the generated result includes the following statistics:

+ n: Number.
+ mean: Mean.
+ sd: Standard deviation.
+ median: Median.
+ trimmed: Trimmed mean with a trim of 0.1.
+ mad: Median absolute deviation (from the median).
+ min: Minimum.
+ max: Maximum.
+ range: The difference between the maximum and minimum value.
+ skew: Skewness.
+ kurtosis: Kurtosis.
+ se: Standard error.


<div class="nav-button left-button">
<a href="imputation.html">← Previous</a>
</div>
<div class="nav-button right-button">
<a href="analysis.html">Next →</a>
</div>

