---
title: "Transformation"
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{transformation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk: 
    collapse: true
    comment: '#>'
    fig.align: "center"
---


## Preliminary


```{r warning=FALSE, message=FALSE, results='hide', fig.show='hide'}
## load R package
library(msDiaLogue)
## preprocessing
fileName <- "../inst/extdata/Toy_Spectronaut_Data.csv"
dataSet <- preprocessing(fileName,
                         filterNaN = TRUE, filterUnique = 2,
                         replaceBlank = TRUE, saveRm = TRUE)
```


## Example


```{r}
dataTran <- transform(dataSet, logFold = 2)
```


<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataTran)
```
</div>


## Details


Raw mass spectrometry intensity measurements are often unsuitable for direct
statistical modeling because the shape of the data is usually not symmetrical
and the variance is not consistent across the range of intensities. Most
proteomic workflows will convert these raw values with a log$_2$ transformation,
which both reshapes the data into a more symmetrical distribution, making it
easier to interpret mean-based fold changes, and also stabilizes the variance
across the intensity range (i.e. reduces heteroscedasticity).


<div class="nav-button left-button">
<a href="preprocessing.html">← Previous</a>
</div>
<div class="nav-button right-button">
<a href="filtering_annotation_based.html">Next →</a>
</div>

