---
title: "A Usage Template for the R Package msDiaLogue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Usage Template}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Load R package

```{r warning=FALSE, message=FALSE}
library(msDiaLogue)
```


## Preprocessing

1. Read the file [Toy_Spectronaut_Data.csv](https://github.com/uconn-scs/msDiaLogue/blob/main/tests/testData/Toy_Spectronaut_Data.csv) into **R**.
2. Omit observations with NaN, and save the removed data in the file
*preprocess_Filtered_Out_NaN.csv* in the current working directory.
3. Filter out proteins with fewer than $2$ unique peptides, and save the removed data
in the file *preprocess_Filtered_Out_Unique.csv* in the current working directory.
4. If there are proteins without names, replace blank protein name entries with their
protein accessions.

```{r results='hide'}
fileName <- "../tests/testData/Toy_Spectronaut_Data.csv"
dataSet <- preprocessing(fileName, filterNaN = TRUE, filterUnique = 2,
                         replaceBlank = TRUE, saveRm = TRUE)
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataSet, align = "l")
```
</div>


## Transformation

Raw intensity measurements, often unsuitable for direct statistical modeling, prompt
transformation in most quantitative proteomics workflow. The preferred solution is
log-transformation, effectively addressing this issue. The log$_2$ transformation is
commonly employed due to its ability to facilitate the straightforward interpretation of
fold changes in protein levels.

```{r}
dataTran <- transform(dataSet, logFold = 2)
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataTran)
```
</div>


## Filtering

For various reasons, such as contamination, lack of informativeness, identification in
only a few samples, etc., data will be filtered before conducting differential analysis to
exclude certain proteins.

For example, the protein 'XPO4_HUMAN' is chosen to be filtered out.

```{r}
dataFilt <- filterOutIn(dataTran, listName = "XPO4_HUMAN",
                        removeList = TRUE, saveRm = TRUE)
```

where `removeList = TRUE` indicates the removal of proteins listed in `listName` from the
`dataTran`. Please note that if `saveRm = TRUE`, the excluded data ("XPO4_HUMAN") will be
saved as a .csv file named *filtered_out_data.csv* in the current working directory.

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataFilt)
```
</div>

Alternatively, if we would to retain proteins like "PA1B2_HUMAN" and "TEBP_HUMAN", simply
set `removelist = FLASE`.

```{r eval=FALSE}
filterOutIn(dataTran, listName = c("PA1B2_HUMAN", "TEBP_HUMAN"),
            removeList = FALSE)
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(filterOutIn(
  dataTran, listName = c("PA1B2_HUMAN", "TEBP_HUMAN"), removeList = FALSE))
```
</div>


## Normalization

Normalization is designed to address systematic biases, enhancing the comparability of
samples while preserving the signal. Various normalization approaches have been proposed.
So far, this package provides three normalization methods for use:

1. "quant": Quantile

2. "median": Column-wise Median

3. "mean": Column-wise Mean

Quantile normalization is generally recommended.

```{r}
dataNorm <- normalize(dataFilt, normalizeType = "quant")
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataNorm)
```
</div>


## Imputation

Mass spectrometry-based quantitative proteomics experiments often yield data with missing
values, potentially impacting downstream analyses. Various imputation methods have been
developed to address this issue. So far, this package provides two imputation methods for
use:

1. "LocalMinVal": It replaces missing values with the lowest value from the protein by
condition combination;

2. "GlobalMinVal": It replaces missing values with the lowest value found within the
entire dataset.

3. "knn": It replaces missing values using the k-nearest neighbors algorithm.

Additional methods will be added later.

Function `dataMissing` summarizes the missingness for each protein.

```{r}
dataMissing <- dataMissing(dataNorm)
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataMissing)
```
</div>

For example, to impute the NA value of `dataNorm` using "LocalMinVal," set the required
percentage of values that must be present in a given protein by condition combination for
values to be imputed to $51\%$.

```{r}
dataImput <- impute(dataNorm, imputeType = "LocalMinVal",
                    reqPercentPresent = 0.51, reportImputing = FALSE)
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataImput)
```
</div>

After the above imputation, there may still be some NA values that consequently need to be
filtered out.

```{r}
dataImput <- filterNA(dataImput, saveRm = TRUE)
```

where `saveRm = TRUE` indicates that the filtered data will be saved as a .csv file named
*filtered_NA_data.csv* in the current working directory.

The `dataImput` is as follows:

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataImput)
```
</div>


## Summarization

Summarization proves valuable in distilling meaningful insights from vast and intricate
data, facilitating the efficient identification, quantification, and interpretation of
protein profiles within complex biological samples.

```{r}
dataSumm <- summarize(dataImput, saveSumm = TRUE)
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataSumm)
```
</div>


## Analysis

### Student's $t$-test

```{r}
anlys_t <- analyze(dataImput, testType = "t-test")
```

Oops! The warning message shows 'Data are essentially constant,' which means that the data
contain proteins with the same value in all samples. In this condition, the P-value of
t-test return NA.

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(anlys_t)
```
</div>


### Volcano

The utility function for `testType = "volcano"` is actually the same as for
`testType = "t-test"`. Here, we remove the proteins "TEBP_HUMAN" and "T126B_HUMAN" to plot
the volcano plot later.

```{r}
anlys_vol <- analyze(dataImput[,colnames(dataImput) != c("TEBP_HUMAN", "T126B_HUMAN")],
                     testType = "volcano")
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(anlys_vol)
```
</div>

### Moderated $t$-test

The main distinction between the Student's and moderated $t$-tests lies in how variance is
computed. While the Student's $t$-test calculates variance based on the data available for
each gene individually, the moderated $t$-est utilizes information from all the chosen
genes to calculate variance.

```{r}
anlys_mod.t <- analyze(dataImput, testType = "mod.t-test")
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(anlys_mod.t)
```
</div>


### MA

```{r}
anlys_MA <- analyze(dataImput, testType = "MA")
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(anlys_MA)
```
</div>


## Visualization

### heatmap

The package offers two options for plotting the heatmap. Option 1 utilizes the source
package `pheatmap`, capable of plotting the dendrogram simultaneously. Option 2 use the
source package `ggplot2` to generates a ggplot object but does not include the dendrogram.

```{r}
visualize(dataImput, graphType = "heatmap",
          pkg = "pheatmap",
          cluster_cols = TRUE, cluster_rows = FALSE,
          show_colnames = TRUE, show_rownames = TRUE)
visualize(dataImput, graphType = "heatmap", pkg = "ggplot2")
```


### MA

```{r}
visualize(anlys_MA, graphType = "MA", transformType = "Log2")
```


### Normalize

```{r}
visualize(dataNorm, graphType = "normalize")
```


### PCA_scree

```{r}
visualize(
  dataImput[,colnames(dataImput) != c("TEBP_HUMAN", "T126B_HUMAN")],
  graphType = "PCA_scree",
  addlabels = TRUE, choice = "variance", ncp = 10)
```


### PCA_ind

```{r}
visualize(
  dataImput[,colnames(dataImput) != c("TEBP_HUMAN", "T126B_HUMAN")],
  graphType = "PCA_ind",
  addlabels = TRUE, addEllipses = TRUE, ellipse.level = 0.95)
```


### PCA_var

```{r}
visualize(
  dataImput[,colnames(dataImput) != c("TEBP_HUMAN", "T126B_HUMAN")],
  graphType = "PCA_var",
  addlabels = TRUE)
```


### PCA_biplot

```{r}
visualize(
  dataImput[,colnames(dataImput) != c("TEBP_HUMAN", "T126B_HUMAN")],
  graphType = "PCA_biplot",
  addEllipses = TRUE, ellipse.level = 0.95, label = "all")
```


### t-test

```{r}
visualize(anlys_mod.t, graphType = "t-test")
```


### Venn

```{r}
dataSort <- sortcondition(dataSet)
visualize(dataSort, graphType = "Venn",
          show_percentage = TRUE,
          fill_color = c("blue", "yellow", "green", "red"),
          show_universal = FALSE)
```


### Volcano

```{r}
visualize(anlys_vol, graphType = "volcano",
          P.thres = 0.05, logF.thres = 0.6)
```


## Other useful function

To observe the data changes of the protein throughout the entire process, this is where
the function `pullProteinPath` comes into play.

```{r}
ZC11B <- pullProteinPath(proteinName = "ZC11B_HUMAN",
                         dataSetList = list(Initial = dataSet,
                                            Transformed = dataTran,
                                            Normalized = dataNorm,
                                            Imputed = dataImput))
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(ZC11B)
```
</div>

