---
title: "A Usage Template for the R Package msDiaLogue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Usage Template}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
```

## Load R package

```{r warning=FALSE, message=FALSE}
library(msDiaLogue)
```


## Preprocessing

**1\.** Load the raw data

+ If the raw data is in a .csv file
[Toy_Spectronaut_Data.csv](https://github.com/uconn-scs/msDiaLogue/blob/main/tests/testData/Toy_Spectronaut_Data.csv),
specify the `fileName` to read the raw data file into **R**.

+ If the raw data is stored as an .RData file
[Toy_Spectronaut_Data.RData](https://github.com/uconn-scs/msDiaLogue/blob/main/tests/testData/Toy_Spectronaut_Data.RData),
first load the data file directly, then specify the `dataSet` in the function.
     
**2\.** Whether to omit observations with NaN

NaN, which stands for 'Not a Number,' represents to protein entries in each condition that
lack a measured quantitative value.

**3\.** The number of unique peptides are required to include a protein

The default is to filter out proteins with fewer than 2 unique peptides. Removing proteins
identified by only 1 unique peptide sequence is more likely to be representatives of the
1% false positive protein identifications. We choose to remove these protein entries from
our downstream analysis in order to provide more confident quantitative results across the
identified proteome. However, these can still be observed in the original protein report
from **Spectronaut**.

**4\.** Whether to rename proteins without names by their accession numbers

**5\.** Whether to save removed data to current working directory

If `saveRm = TRUE`, the data removed in step 2 (*preprocess_Filtered_Out_NaN.csv*) and
step 3 (*preprocess_Filtered_Out_Unique.csv*) will be saved in the current working
directory.

```{r eval=FALSE}
## if the raw data is in a .csv file
fileName <- "../tests/testData/Toy_Spectronaut_Data.csv"
dataSet <- preprocessing(fileName,
                         filterNaN = TRUE, filterUnique = 2,
                         replaceBlank = TRUE, saveRm = TRUE)
```
```{r results='hide'}
## if the raw data is in an .Rdata file
load("../tests/testData/Toy_Spectronaut_Data.RData")
dataSet <- preprocessing(dataSet = Toy_Spectronaut_Data,
                         filterNaN = TRUE, filterUnique = 2,
                         replaceBlank = TRUE, saveRm = TRUE)
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataSet, align = "l")
```
</div>


## Transformation

Raw intensity measurements, often unsuitable for direct statistical modeling, prompt
transformation in most quantitative proteomics workflow. The preferred solution is
log-transformation, effectively addressing this issue. The log$_2$ transformation is
commonly employed due to its ability to facilitate the straightforward interpretation of
fold changes in protein levels.

```{r}
dataTran <- transform(dataSet, logFold = 2)
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataTran)
```
</div>


## Filtering

For various reasons, such as the presence of high-abundance contaminant proteins that may
be irrelevant to the present analysis, a lack of informativeness (e.g., proteins that
exhibit no differential abundance at all or proteins that are known with certainty to be
false positive identifications with more than one unique peptide identified),
identification of a protein exclusively in a small subset of samples or only one
experimental condition, etc., we recommend data filtering before conducting differential
analysis. The application of protein-level filters is optional and is expected to vary by
project.

**Case 1.** Remove the listed proteins and retain the remaining proteins

For example, the protein "XPO4_HUMAN" is chosen to be filtered out.

```{r eval=FALSE}
filterOutIn(dataTran, listName = "XPO4_HUMAN",
            removeList = TRUE, saveRm = TRUE)
```

where `removeList = TRUE` indicates the removal of proteins listed in `listName` from the
`dataTran`. Please note that if `saveRm = TRUE`, the excluded data ("XPO4_HUMAN") will be
saved as a .csv file named *filtered_out_data.csv* in the current working directory.

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(
  filterOutIn(dataTran, listName = "XPO4_HUMAN", removeList = TRUE, saveRm = TRUE))
```
</div>

**Case 2.** Keep the listed proteins and remove the rest.

Alternatively, if we would to keep proteins like "PA1B2_HUMAN" and "TEBP_HUMAN", simply
set `removelist = FALSE`.

```{r eval=FALSE}
filterOutIn(dataTran, listName = c("PA1B2_HUMAN", "TEBP_HUMAN"),
            removeList = FALSE)
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(filterOutIn(
  dataTran, listName = c("PA1B2_HUMAN", "TEBP_HUMAN"), removeList = FALSE))
```
</div>


## Normalization

Normalization is designed to address systematic biases, enhancing the comparability of
samples while preserving the signal. Various normalization approaches have been proposed.
So far, this package provides three normalization methods for use:

1. "quant": Quantile

2. "median": Column-wise Median

3. "mean": Column-wise Mean

Quantile normalization is generally recommended.

```{r}
dataNorm <- normalize(dataTran, normalizeType = "quant")
```

Oh! The message "Warning: Removed 16 rows containing non-finite values" indicates the
presence of 16 NA (Not Available) values in the data. These NA values are are
automatically excluded when generating the boxplot but retained in the actual dataset.

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataNorm)
```
</div>


## Imputation

Traditional Data-Dependent Acquisition (DDA) mass spectrometry-based quantitative
proteomics datasets often yield many missing quantitative values due to limits in sampling
frequency. Data-Independent Acquisition (DIA) helps to substantially reduce the number of
missing values compared to DDA experiments, but some will still exist due to sample
variability, and these missing values will impact downstream analyses. Various imputation
methods have been developed to address this issue and assign a quantitative value to
proteins with missing quantitative values. So far, this package provides five imputation
methods for use:

1. "LocalMinVal": It replaces missing values with the lowest value from the protein by
condition combination;

2. "GlobalMinVal": It replaces missing values with the lowest value found within the
entire dataset.

3. "knn": It replaces missing values using the k-nearest neighbors algorithm.

4. "seq-knn": It replaces missing values using the sequential k-nearest neighbors algorithm.

5. "trunc-knn": It replaces missing values using the truncated k-nearest neighbors algorithm.

Additional methods will be added later.

Function `dataMissing` summarizes the missingness for each protein,

```{r}
dataMissing <- dataMissing(dataNorm, show_labels = TRUE)
```

where `show_labels = TRUE` means that the protein names are displayed in the plot. The
percentage in the protein labels represents the proportion of missing data in the samples
for that protein. For instance, the label "XPO4_HUMAN (80\%)" indicates that, within all
observations for the protein "XPO4_HUMAN", 80\% of the data is missing. Additionally, the
percentage in the legend represents the proportion of missing data in the whole dataset.
Here, 8.4\% of the data in `dataNorm` is missing.

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataMissing)
```
</div>

In addition, the table shows the number and percentage of missing data for every protein.
For example, for the protein "XPO4_HUMAN," there are 8 NA values in the samples,
constituting 50\% of the missing data in the entire dataset.

For example, to impute the NA value of `dataNorm` using "LocalMinVal," set the required
percentage of values that must be present in a given protein by condition combination for
values to be imputed to $51\%$.

```{r}
dataImput <- impute(dataNorm, imputeType = "LocalMinVal",
                    reqPercentPresent = 0.51, reportImputing = FALSE)
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataImput)
```
</div>

After the above imputation, there may still be some NA values that consequently need to be
filtered out.

```{r}
dataImput <- filterNA(dataImput, saveRm = TRUE)
```

where `saveRm = TRUE` indicates that the filtered data will be saved as a .csv file named
*filtered_NA_data.csv* in the current working directory.

The `dataImput` is as follows:

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataImput)
```
</div>


## Summarization

Summarization proves valuable in distilling meaningful insights from vast and intricate
data, facilitating the efficient identification, quantification, and interpretation of
protein profiles within complex biological samples.

```{r}
dataSumm <- summarize(dataImput, saveSumm = TRUE)
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(dataSumm)
```
</div>


## Analysis

### Student's $t$-test

```{r}
anlys_t <- analyze(dataImput, testType = "t-test")
```

Oops! The warning message shows 'Data are essentially constant,' which means that the data
contain proteins with the same value in all samples. In this condition, the P-value of
t-test return NA.

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(anlys_t)
```
</div>


### Volcano

The utility function for `testType = "volcano"` is actually the same as for
`testType = "t-test"`. Here, we remove the proteins "TEBP_HUMAN" and "T126B_HUMAN" to plot
the volcano plot later.

```{r}
anlys_vol <- analyze(dataImput[,colnames(dataImput) != c("TEBP_HUMAN", "T126B_HUMAN")],
                     testType = "volcano")
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(anlys_vol)
```
</div>

### Moderated $t$-test

The main distinction between the Student's and moderated $t$-tests lies in how variance is
computed. While the Student's $t$-test calculates variance based on the data available for
each gene individually, the moderated $t$-est utilizes information from all the chosen
genes to calculate variance.

```{r}
anlys_mod.t <- analyze(dataImput, testType = "mod.t-test")
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(anlys_mod.t)
```
</div>


### MA

```{r}
anlys_MA <- analyze(dataImput, testType = "MA")
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(anlys_MA)
```
</div>


## Visualization

### heatmap

The package offers two options for plotting the heatmap. Option 1 utilizes the source
package `pheatmap`, capable of plotting the dendrogram simultaneously. Option 2 use the
source package `ggplot2` to generates a ggplot object but does not include the dendrogram.

```{r}
visualize(dataImput, graphType = "heatmap",
          pkg = "pheatmap",
          cluster_cols = TRUE, cluster_rows = FALSE,
          show_colnames = TRUE, show_rownames = TRUE)
visualize(dataImput, graphType = "heatmap", pkg = "ggplot2")
```


### MA

```{r}
visualize(anlys_MA, graphType = "MA", transformType = "Log2")
```


### Normalize

```{r}
visualize(dataNorm, graphType = "normalize")
```


### PCA_scree

```{r}
visualize(
  dataImput[,colnames(dataImput) != c("TEBP_HUMAN", "T126B_HUMAN")],
  graphType = "PCA_scree",
  addlabels = TRUE, choice = "variance", ncp = 10)
```


### PCA_ind

```{r}
visualize(
  dataImput[,colnames(dataImput) != c("TEBP_HUMAN", "T126B_HUMAN")],
  graphType = "PCA_ind",
  addlabels = TRUE, addEllipses = TRUE, ellipse.level = 0.95)
```


### PCA_var

```{r}
visualize(
  dataImput[,colnames(dataImput) != c("TEBP_HUMAN", "T126B_HUMAN")],
  graphType = "PCA_var",
  addlabels = TRUE)
```


### PCA_biplot

```{r}
visualize(
  dataImput[,colnames(dataImput) != c("TEBP_HUMAN", "T126B_HUMAN")],
  graphType = "PCA_biplot",
  addEllipses = TRUE, ellipse.level = 0.95, label = "all")
```


### t-test

```{r}
visualize(anlys_mod.t, graphType = "t-test")
```

### Upset

```{r}
dataSort <- sortcondition(dataSet)
visualize(dataSort, graphType = "Upset")
```

### Venn

```{r}
visualize(dataSort, graphType = "Venn",
          show_percentage = TRUE,
          fill_color = c("blue", "yellow", "green", "red"),
          show_universal = FALSE)
```


### Volcano

```{r}
visualize(anlys_vol, graphType = "volcano",
          P.thres = 0.05, logF.thres = 0.6)
```


## Other useful function

To observe the data changes of the protein throughout the entire process, this is where
the function `pullProteinPath` comes into play.

```{r}
ZC11B <- pullProteinPath(proteinName = "ZC11B_HUMAN",
                         dataSetList = list(Initial = dataSet,
                                            Transformed = dataTran,
                                            Normalized = dataNorm,
                                            Imputed = dataImput))
```

<div style="overflow-x: auto;">
```{r echo=FALSE}
knitr::kable(ZC11B)
```
</div>

